preciseTAD: A machine learning framework for precise TAD boundary prediction at base-level resolution

### 2020-08-25 Manuscript/manuscript.Rmd
Major edits

**Background.** High-throughput chromosome conformation capture technology (Hi-C) revealed extensive DNA looping and folding into discrete 3D domains. These include Topologically Associated Domains (TADs) and chromatin loops, 3D domains critical for cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. However, the high resolution of genomic annotations associated with boundaries, such as CTCF, members of cohesin complex, suggests they can inform the precise location of domain boundaries. Several methods attempted to leverage genome annotation data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary, and a much smaller number of boundaries than the rest of the genome (class imbalance).

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. <!--Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

Cell-line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

We assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture, including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell-type-specific genome annotation data.

@@

A total of 77 cell-line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentation, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Supplementary Figure 1).

@@

![**A diagram of the machine learning framework used to build TAD/loop boundary region prediction models.** Step 1 defines the dimensions of the data matrix $A_{N \times (p+1)}$ used for training and testing by establishing a resolution, $r$, at which to bin the linear genome into $N$ genomic bins for each chromosome, except a holdout chromosome *i*. The response vector **Y** is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). A feature engineering procedure is selected to define the feature space of $A$ composed of $p$ genomic annotations. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a resampling technique to the training data to address the class imbalance. Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8,052                       |                             15,468                           |                535,363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7,676                       |                             14,253                           |                267,682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4,670                       |                             8,363                            |                107,073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2,349                       |                             4,224                            |                 53,537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1,031                       |                             1,883                            |                 26,768               |           0.07         |
|        Peakachu       |       10   kb     |                        16,185                       |                          21,421                              |                267,682               |        0.14            |
Table: TAD/loop boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

## Transcription factor binding sites outperform histone- and chromatin segmentation-specific models

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin segmentation, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each chromosome set. To determine transcription factors most important for boundary prediction, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). When training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on Peakachu resulted in an average AUC=0.885 and slightly reduced to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). These results suggest that a classifier trained on the data from one cell line performs well when using data from another cell line.

![**Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and subsequent area under the curves (AUCs) when training and testing on the same cell line versus across cell lines. The curves represent the average sensitivities and specificities across each holdout chromosome.](Figures/figure4.png)

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with the corresponding genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around medoids clustering (CLARA) to predict the boundary location at base resolution (see Methods).

@@

Next, we evaluated the biological significance of *preciseTAD* boundaries. Peak regions associated with known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143, were found to co-localize more frequently around *preciseTAD* predicted TAD and loop boundaries compared to Arrowhead called TAD and Peakached loop boundaries respectively (Figure 5B-5C; Supplementary Figure 5). Furthermore, signal heatmaps indicated that our predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Supplementary Figures 6-7). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top-ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 6A, Supplementary Figure 8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of ARROWHEAD called TAD boundaries (blue) vs *preciseTAD* predicted TAD boundaries (green) on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. Genome browser visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 is presented underneath. Profile plots comparing the signal strength of top TFBS around flanked (B) ARROWHEAD called TAD boundaries (blue) and *preciseTAD* predicted TAD boundaries (green), and (C) around Peakachu loop boundaries (red) and *preciseTAD* predicted loop boundaries (green).](Figures/figure5.png)

@@

TAD boundaries are relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 6B). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

![**preciseTAD predicted boundaries are both statistically significantly closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distances from called and predicted boundaries to nearest CTCF sites. The p-value is derived from the Wilcoxon Rank Sum test. (B) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by *preciseTAD* (using ARROWHEAD; top left), ARROWHEAD (top right), *preciseTAD* (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram.](Figures/figure6.png)

## *preciseTAD* predicted TAD boundaries are nested within predicted loop boundaries

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop domain boundaries may be driven by similar molecular mechanisms.

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution TAD/loop boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

>>>>

**Background.** High-throughput chromosome conformation capture technology (Hi-C) revealed extensive DNA looping and folding into discrete 3D domains. These include Topologically Associating Domains (TADs) and chromatin loops, 3D domains critical for cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. However, the high resolution of genomic annotations associated with boundaries, such as CTCF, members of cohesin complex, suggests they can inform the precise location of domain boundaries. Several methods attempted to leverage genome annotation data for predicting domain boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary, and a much smaller number of boundaries than the rest of the genome (class imbalance).

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associating Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. <!--Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state data (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin states (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

Cell-line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin states lack signal values.

@@

We assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture, including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy] [??? Correct, change to new pipeline]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell-type-specific genome annotation data.

@@

A total of 77 cell-line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Supplementary Figure 1).

@@

![**A diagram of the machine learning framework used to build domain boundary region prediction models.** Step 1 defines the dimensions of the data matrix $A_{N \times (p+1)}$ used for training and testing by establishing a resolution, $r$, at which to bin the linear genome into $N$ genomic bins for each chromosome, except a holdout chromosome *i*. The response vector **Y** is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). A feature engineering procedure is selected to define the feature space of $A$ composed of $p$ genomic annotations. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a resampling technique to the training data to address the class imbalance. Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8,052                       |                             15,468                           |                535,363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7,676                       |                             14,253                           |                267,682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4,670                       |                             8,363                            |                107,073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2,349                       |                             4,224                            |                 53,537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1,031                       |                             1,883                            |                 26,768               |           0.07         |
|        Peakachu       |       10   kb     |                        16,185                       |                          21,421                              |                267,682               |        0.14            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

## Transcription factor binding sites outperform histone- and chromatin state-specific models

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each chromosome set. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from GM12878 cell line resulted in an average AUC=0.845 (Figure 4D). The performance of this model only slightly decreased when evaluated using K562-specific Arrowhead boundaries and genomic annotation data (AUC = 0.803). Likewise, the performance of the model trained using Peakachu boundaries and genomic annotation data from GM12878 cell line was comparable when evaluated using K562-specific Peakachu boundaries and genomic annotations (AUC=0.885 and 0.868, respectively). These results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). In summary, our observations suggest that a classifier trained on the data from one cell line performs well when using data from another cell line.

![**Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and subsequent area under the curves (AUCs) when training and testing on the same cell line versus across cell lines. The curves represent the average sensitivities and specificities across each holdout chromosome.](Figures/figure4.png)

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with the corresponding genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around medoids clustering (CLARA) to predict the location of boundary regions and summit points (highest boundary probability) at base resolution (see Methods).

@@

Next, we evaluated the biological significance of *preciseTAD* boundaries. Peak regions associated with known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143, co-localized more frequently around *preciseTAD*-predicted TAD and loop boundaries compared to Arrowhead called TAD and Peakached loop boundaries respectively (Figure 5B-5C; Supplementary Figure 5). Furthermore, signal heatmaps indicated that our predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Supplementary Figures 6-7). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Supplementary Figure 8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B, C) Profile plots comparing the signal strength of CTCF, RAD21, SMC3, and ZNF143 around Arrowhead-called boundaries (blue, B), Peakachu loop boundaries (red, C) vs. *preciseTAD*-predicted boundaries (green).](Figures/figure5.png)

@@

TAD boundaries are relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35%/55% of *preciseTAD*-predicted domain boundaries using models trained on Arrowhead/Peakachu data, respectively, were conserved between GM12878 and K562 cell lines. This is in contrast to only 26% and 55% for Arrowhead and Peakachu boundaries themselves (Figure 6B). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

![**preciseTAD-predicted boundaries are both statistically significantly closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distances from called and predicted boundaries to nearest CTCF sites. P-value are from the Wilcoxon Rank Sum test. (B) Venn diagrams illustrating the levels of conservation between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead-trained *preciseTAD* model (top left), Arrowhead boundaries themselves (top right), Peakachu-trained *preciseTAD* model (bottom left), and Peakachu boundaries themselves (bottom right). J - the median Jaccard Index (IQR in parentheses) across all chromosomes.](Figures/figure6.png)

## *preciseTAD*-predicted TAD boundaries are nested within predicted loop boundaries

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.
